---
title: "Decision Tree Statistical Analyses in R"
subtitle: "Example R code linked to the prepared datasets"
author: "Darias Holgado & Corinna Martarelli"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)

data_dir <- file.path("..", "data", "examples")

read_example <- function(file) {
  read.csv(file.path(data_dir, file), stringsAsFactors = FALSE)
}

has_pkg <- function(package) {
  requireNamespace(package, quietly = TRUE)
}

optional_packages <- c(
  "ARTool",
  "caret",
  "e1071",
  "lavaan",
  "lme4",
  "lmerTest",
  "MASS",
  "metafor",
  "nnet",
  "ordinal",
  "performance",
  "psych",
  "randomForest",
  "rpart"
)

install_missing_packages <- function(packages = optional_packages) {
  missing_packages <- packages[!vapply(packages, has_pkg, logical(1))]

  if (length(missing_packages) == 0) {
    message("All optional packages are already installed.")
    return(invisible(character()))
  }

  install.packages(
    missing_packages,
    repos = "https://cloud.r-project.org"
  )

  invisible(missing_packages)
}
```

## Optional packages

Run this chunk once if you want to install only the packages that are missing on your computer.

```{r install-missing-packages, eval=FALSE}
install_missing_packages()
```

# Associations and relationships

## Pearson correlation

```{r pearson-correlation}
data <- read_example("pearson.csv")
cor.test(data$x, data$y, method = "pearson")
plot(data$x, data$y, xlab = "x", ylab = "y")
```

## Spearman rank correlation

```{r spearman-correlation}
data <- read_example("spearman.csv")
cor.test(data$stress_rank, data$wellbeing_rank, method = "spearman", exact = FALSE)
```

## Kendall rank correlation

```{r kendall-correlation}
data <- read_example("kendall.csv")
cor.test(data$adherence_rank, data$wellbeing_rank, method = "kendall", exact = FALSE)
```

## Chi-square test of association

```{r chi-square-association}
data <- read_example("chiSquareAssociation.csv")
tab <- table(data$group, data$outcome)
tab
chisq.test(tab)
```

## Fisher exact test

```{r fisher-exact}
data <- read_example("fisher.csv")
tab <- table(data$treatment, data$response)
tab
fisher.test(tab)
```

## McNemar test

```{r mcnemar-test}
data <- read_example("mcnemar.csv")
tab <- table(data$before, data$after)
tab
mcnemar.test(tab)
```

## Log-linear model

```{r log-linear-model}
data <- read_example("logLinearModel.csv")
tab <- xtabs(count ~ therapy + outcome + gender, data = data)

if (has_pkg("MASS")) {
  fit <- MASS::loglm(~ therapy * outcome + gender, data = tab)
  summary(fit)
} else {
  message("Install MASS to run MASS::loglm().")
}
```

## Path analysis / mediation

```{r path-analysis}
data <- read_example("pathAnalysis.csv")

if (has_pkg("lavaan")) {
  model <- '
    mediator ~ a * predictor
    outcome ~ b * mediator + c_prime * predictor
    indirect := a * b
    total := c_prime + indirect
  '

  fit <- lavaan::sem(model, data = data, se = "bootstrap", bootstrap = 5000)
  summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE,ci = TRUE)
} else {
  message("Install lavaan to run the mediation model.")
}

#Extract path coefficients

lavaan::parameterEstimates(
  fit,
  standardized = TRUE,
  ci = TRUE
)

#Extract regression path
pe <- lavaan::parameterEstimates(
  fit,
  standardized = TRUE,
  ci = TRUE
)

pe[pe$op == "~", ]

#R2

lavaan::inspect(fit, "r2")

```

## Structural equation modeling

```{r structural-equation-modeling}
data <- read_example("structuralEquationModeling.csv")

if (has_pkg("lavaan")) {
  model <- '
    stress =~ stress1 + stress2 + stress3
    recovery =~ sleep1 + sleep2 + sleep3
    recovery ~ stress
    performance ~ recovery + stress
  '

  fit <- lavaan::sem(model, data = data)
  summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
} else {
  message("Install lavaan to run SEM.")
}

pe <- lavaan::parameterEstimates(
  fit,
  standardized = TRUE,
  ci = TRUE
)

pe


```

## Discriminant analysis

```{r discriminant-analysis}
data <- read_example("discriminantAnalysis.csv")
data$group <- factor(data$group)

if (has_pkg("MASS")) {
  fit <- MASS::lda(group ~ anxiety + motivation + reaction_time, data = data)
  fit
  pred <- predict(fit)$class
  table(Predicted = pred, Observed = data$group)
  mean(pred == data$group)
} else {
  message("Install MASS to run linear discriminant analysis.")
}
```

# One-sample analyses

## One-sample t-test

```{r one-sample-t-test}
data <- read_example("oneSampleT.csv")
t.test(data$score, mu = 0)
```

## One-sample z-test

```{r one-sample-z-test}
data <- read_example("oneSampleZ.csv")

mu0 <- 50
sigma <- 15
x <- data$score
n <- length(x)
z <- (mean(x) - mu0) / (sigma / sqrt(n))
p <- 2 * pnorm(-abs(z))

c(
  sample_mean = mean(x),
  null_mean = mu0,
  population_sd = sigma,
  n = n,
  z = z,
  p = p
)
```

## One-sample Wilcoxon signed-rank test

```{r one-sample-wilcoxon}
data <- read_example("oneSampleWilcoxon.csv")
wilcox.test(data$rating, mu = 0)
```

# Two-group and paired analyses

## Independent-samples t-test

```{r independent-t-test}
data <- read_example("independentT.csv")
data$group <- factor(data$group)
t.test(score ~ group, data = data, var.equal = FALSE)
```

## Mann-Whitney U test

```{r mann-whitney-u}
data <- read_example("mannWhitney.csv")
data$group <- factor(data$group)
wilcox.test(pain_score ~ group, data = data)
```

## Paired-samples t-test

```{r paired-t-test}
data <- read_example("pairedT.csv")
t.test(data$baseline_score, data$followup_score, paired = TRUE)
shapiro.test(data$difference)
```

## Wilcoxon signed-rank test

```{r wilcoxon-signed-rank}
data <- read_example("wilcoxon.csv")
wilcox.test(data$before, data$after, paired = TRUE)
```

# Variance tests

## Chi-square test for one variance

```{r chi-square-variance}
data <- read_example("chiSquareVariance.csv")

n <- length(data$score)
s2 <- var(data$score)
sigma2_0 <- 100
chi2 <- (n - 1) * s2 / sigma2_0
p <- 2 * min(pchisq(chi2, df = n - 1), 1 - pchisq(chi2, df = n - 1))

c(
  chi_square = chi2,
  df = n - 1,
  p = p
)
```

## F-test for two variances

```{r f-test-variance}
data <- read_example("varianceFTest.csv")
data$group <- factor(data$group)
var.test(score ~ group, data = data)
```

# ANOVA and nonparametric alternatives

## One-way ANOVA

```{r one-way-anova}
data <- read_example("anova.csv")
data$group <- factor(data$group)

fit <- aov(score ~ group, data = data)
summary(fit)
TukeyHSD(fit)
```

## Repeated-measures ANOVA

```{r repeated-measures-anova}
data <- read_example("repeatedAnova.csv")

long_data <- data.frame(
  id = rep(data$id, each = 3),
  condition = factor(rep(c("baseline", "week4", "week8"), times = nrow(data))),
  score = c(t(data[, c("baseline", "week4", "week8")]))
)

fit <- aov(score ~ condition + Error(id / condition), data = long_data)
summary(fit)
```

## Two-way ANOVA

```{r two-way-anova}
data <- read_example("twoWayAnova.csv")
data$therapy <- factor(data$therapy)
data$age_group <- factor(data$age_group)

fit <- aov(stress_score ~ therapy * age_group, data = data)
summary(fit)
```

## Two-way repeated-measures ANOVA

```{r two-way-repeated-anova}
data <- read_example("twoWayRepeatedAnova.csv")

long_data <- data.frame(
  participant = rep(data$participant, each = 6),
  load = factor(rep(rep(c("low", "high"), each = 3), times = nrow(data))),
  time = factor(rep(c("pre", "post", "followup"), times = 2 * nrow(data))),
  score = c(t(data[, c("low_pre", "low_post", "low_followup", "high_pre", "high_post", "high_followup")]))
)

fit <- aov(score ~ load * time + Error(participant / (load * time)), data = long_data)
summary(fit)
```

## Kruskal-Wallis test

```{r kruskal-wallis}
data <- read_example("kruskalWallis.csv")
data$group <- factor(data$group)
kruskal.test(rank_score ~ group, data = data)
```

## Friedman test

```{r friedman-test}
data <- read_example("friedman.csv")
friedman.test(as.matrix(data[, c("condition_a", "condition_b", "condition_c")]))
```

## Nonparametric two-way ANOVA via aligned rank transform

```{r nonparametric-two-way-anova}
data <- read_example("nonparametricTwoWayAnova.csv")
data$therapy <- factor(data$therapy)
data$age_group <- factor(data$age_group)

if (has_pkg("ARTool")) {
  fit <- ARTool::art(stress_rating ~ therapy * age_group, data = data)
  anova(fit)
} else {
  message("Install ARTool to run aligned rank transform ANOVA.")
}
```

## Nonparametric two-way repeated-measures ANOVA via aligned rank transform

```{r nonparametric-two-way-repeated-anova}
data <- read_example("nonparametricTwoWayRepeatedAnova.csv")

long_data <- data.frame(
  participant = factor(rep(data$participant, each = 6)),
  load = factor(rep(rep(c("low", "high"), each = 3), times = nrow(data))),
  time = factor(rep(c("pre", "post", "followup"), times = 2 * nrow(data))),
  stress_rating = c(t(data[, c("low_pre", "low_post", "low_followup", "high_pre", "high_post", "high_followup")]))
)

if (has_pkg("ARTool")) {
  fit <- ARTool::art(stress_rating ~ load * time + Error(participant), data = long_data)
  anova(fit)
} else {
  message("Install ARTool to run aligned rank transform ANOVA.")
}
```

# Frequencies and distributions

## Binomial test

```{r binomial-test}
data <- read_example("binomialTest.csv")
successes <- sum(data$response == "yes")
n <- length(data$response)
binom.test(successes, n, p = 0.50)
```

## Chi-square goodness-of-fit test

```{r chi-square-goodness}
data <- read_example("chiSquareGoodness.csv")
observed <- table(data$preference)
observed
chisq.test(observed, p = rep(1 / length(observed), length(observed)))
```

# Regression models

Regression models estimate associations and predictions. They do not, by themselves, provide evidence of causality.

## Linear regression

```{r linear-regression}
data <- read_example("linearRegression.csv")

fit <- lm(
  race_time_min ~ training_volume_km_week + vo2max_ml_kg_min + age_years,
  data = data
)

summary(fit)
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

## Binary logistic regression

```{r binary-logistic-regression}
data <- read_example("logisticRegression.csv")
data$injury <- factor(data$injury, levels = c("no", "yes"))

fit <- glm(
  injury ~ training_load_au + previous_injury + recovery_score + age_years,
  data = data,
  family = binomial
)

summary(fit)
exp(coef(fit))
```

## Multinomial logistic regression

```{r multinomial-logistic-regression}
data <- read_example("multinomialRegression.csv")

if (has_pkg("nnet")) {
  data$treatment_outcome <- relevel(factor(data$treatment_outcome), ref = "completion")
  fit <- nnet::multinom(
    treatment_outcome ~ baseline_symptom_score + motivation_score + previous_therapy + age_years,
    data = data
  )
  summary(fit)
} else {
  message("Install nnet to run multinomial logistic regression.")
}
```

## Linear mixed model

```{r linear-mixed-model}
data <- read_example("linearMixedModel.csv")
data$participant <- factor(data$participant)
data$condition <- factor(data$condition)
data$team <- factor(data$team)

if (has_pkg("lme4") && has_pkg("lmerTest")) {
  fit <- lmerTest::lmer(score ~ condition + week + (1 | participant), data = data)
  summary(fit)

  if (has_pkg("performance")) {
    performance::r2_nakagawa(fit)
  }
} else {
  message("Install lme4 and lmerTest to run the linear mixed model.")
}
```

## Generalized linear mixed model

```{r generalized-linear-mixed-model}
data <- read_example("generalizedMixedModel.csv")
data$participant <- factor(data$participant)
data$condition <- factor(data$condition)
data$success_num <- ifelse(data$success %in% c("yes", "1", 1, "success"), 1, 0)

if (has_pkg("lme4")) {
  fit <- lme4::glmer(
    success_num ~ condition + week + (1 | participant),
    data = data,
    family = binomial
  )
  summary(fit)
  exp(lme4::fixef(fit))
} else {
  message("Install lme4 to run the generalized linear mixed model.")
}
```

## Ordinal mixed model

```{r ordinal-mixed-model}
data <- read_example("ordinalMixedModel.csv")
data$participant <- factor(data$participant)
data$condition <- factor(data$condition)
data$rating <- ordered(data$rating)

if (has_pkg("ordinal")) {
  fit <- ordinal::clmm(rating ~ condition + day + (1 | participant), data = data)
  summary(fit)
  exp(coef(fit))
} else {
  message("Install ordinal to run the ordinal mixed model.")
}
```

# Exploratory structure

## Factor analysis

```{r factor-analysis}
data <- read_example("factorAnalysis.csv")
items <- c(
  "stress_item_1", "stress_item_2", "stress_item_3",
  "mood_item_1", "mood_item_2", "mood_item_3",
  "sleep_item_1", "sleep_item_2"
)

if (has_pkg("psych")) {
  psych::fa.parallel(data[, items], fa = "fa")
  fit <- psych::fa(data[, items], nfactors = 3, rotate = "oblimin")
  print(fit$loadings, cutoff = .30)
} else {
  fit <- factanal(data[, items], factors = 3, rotation = "promax")
  print(fit)
}
```

## Cluster analysis

```{r cluster-analysis}
data <- read_example("clusterAnalysis.csv")
variables <- c("training_load", "recovery_score", "sprint_time", "jump_height", "injury_count")

scaled <- scale(data[, variables])
d <- dist(scaled, method = "euclidean")
fit <- hclust(d, method = "ward.D2")
plot(fit)

clusters <- cutree(fit, k = 3)
table(clusters)

set.seed(1)
kmeans_fit <- kmeans(scaled, centers = 3, nstart = 25)
table(kmeans_fit$cluster)
```

## Multidimensional scaling

```{r multidimensional-scaling}
data <- read_example("multidimensionalScaling.csv")
variables <- c("calm", "joy", "anger", "sadness", "fear")

d <- dist(scale(data[, variables]))
fit <- cmdscale(d, k = 2, eig = TRUE)

plot(fit$points, xlab = "Dimension 1", ylab = "Dimension 2")
text(fit$points, labels = data$object, pos = 3)

if (has_pkg("MASS")) {
  nonmetric_fit <- MASS::isoMDS(d, k = 2)
  nonmetric_fit$stress
}
```

# Meta-analysis

## Odds ratio meta-analysis

```{r meta-odds-ratio}
data <- read_example("metaOddsRatio.csv")

if (has_pkg("metafor")) {
  data$non_events_treatment <- data$total_treatment - data$events_treatment
  data$non_events_control <- data$total_control - data$events_control

  es <- metafor::escalc(
    measure = "OR",
    ai = events_treatment,
    bi = non_events_treatment,
    ci = events_control,
    di = non_events_control,
    data = data
  )

  fit <- metafor::rma(yi, vi, mods = ~ moderator, data = es, method = "REML")
  summary(fit)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Correlation coefficient meta-analysis

```{r meta-correlation}
data <- read_example("metaCorrelation.csv")

if (has_pkg("metafor")) {
  es <- metafor::escalc(measure = "ZCOR", ri = correlation, ni = sample_size, data = data)
  fit <- metafor::rma(yi, vi, mods = ~ moderator, data = es, method = "REML")
  summary(fit)
  predict(fit, transf = metafor::transf.ztor)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Mean difference meta-analysis

```{r meta-mean-difference}
data <- read_example("metaMeanDifferences.csv")

if (has_pkg("metafor")) {
  es <- metafor::escalc(
    measure = "SMD",
    m1i = mean_treatment,
    sd1i = sd_treatment,
    n1i = n_treatment,
    m2i = mean_control,
    sd2i = sd_control,
    n2i = n_control,
    data = data
  )

  fit <- metafor::rma(yi, vi, mods = ~ moderator, data = es, method = "REML")
  summary(fit)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Effect size meta-analysis

```{r meta-effect-size}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    vi = variance,
    mods = ~ moderator,
    data = data,
    method = "REML"
  )

  summary(fit)
  forest(fit, slab = data$study_label)
  funnel(fit)
  metafor::regtest(fit, model = "rma")
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Fixed-effect meta-analysis

```{r fixed-effect-meta-analysis}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    sei = se,
    data = data,
    method = "FE"
  )

  summary(fit)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Random-effects meta-analysis

```{r random-effects-meta-analysis}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    sei = se,
    data = data,
    method = "REML"
  )

  summary(fit)
  predict(fit)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Subgroup meta-analysis

```{r subgroup-meta-analysis}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    sei = se,
    mods = ~ moderator,
    data = data,
    method = "REML"
  )

  summary(fit)

  by(
    data,
    data$moderator,
    function(d) summary(metafor::rma(yi = effect_size, sei = se, data = d, method = "REML"))
  )
} else {
  message("Install metafor to run this meta-analysis.")
}
```

## Meta-regression

```{r meta-regression}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    sei = se,
    mods = ~ moderator,
    data = data,
    method = "REML"
  )

  summary(fit)
} else {
  message("Install metafor to run meta-regression.")
}
```

## Publication-bias diagnostics

```{r publication-bias-diagnostics}
data <- read_example("metaEffectSizes.csv")

if (has_pkg("metafor")) {
  fit <- metafor::rma(
    yi = effect_size,
    sei = se,
    data = data,
    method = "REML"
  )

  funnel(fit)
  metafor::regtest(fit, model = "rma")
  metafor::trimfill(fit)
} else {
  message("Install metafor to run publication-bias diagnostics.")
}
```

## Proportion meta-analysis

```{r meta-proportion}
data <- read_example("metaProportion.csv")

if (has_pkg("metafor")) {
  es <- metafor::escalc(
    measure = "PLO",
    xi = event_frequency,
    ni = total_sample_size,
    data = data
  )

  fit <- metafor::rma(yi, vi, mods = ~ moderator, data = es, method = "REML")
  summary(fit)
  predict(fit, transf = metafor::transf.ilogit)
  forest(fit, slab = data$study_label)
} else {
  message("Install metafor to run this meta-analysis.")
}
```

# Machine learning regression

## Decision tree regression

```{r decision-tree-regression}
data <- read_example("decisionTreeRegression.csv")
model_data <- subset(data, select = -client)

if (has_pkg("rpart")) {
  set.seed(1)
  train_id <- sample(seq_len(nrow(model_data)), size = floor(0.70 * nrow(model_data)))

  fit <- rpart::rpart(
    outcome_symptom_score ~ motivation_score + baseline_distress_score + sessions_attended,
    data = model_data[train_id, ],
    method = "anova",
    control = rpart::rpart.control(cp = .01)
  )

  pred <- predict(fit, newdata = model_data[-train_id, ])
  actual <- model_data$outcome_symptom_score[-train_id]
  c(
    RMSE = sqrt(mean((pred - actual)^2)),
    MAE = mean(abs(pred - actual)),
    R2 = cor(pred, actual)^2
  )
  plot(fit)
  text(fit)
} else {
  message("Install rpart to run decision tree regression.")
}
```

## Random forest regression

```{r random-forest-regression}
data <- read_example("randomForestRegression.csv")
model_data <- subset(data, select = -athlete)

if (has_pkg("randomForest")) {
  set.seed(1)
  train_id <- sample(seq_len(nrow(model_data)), size = floor(0.70 * nrow(model_data)))

  fit <- randomForest::randomForest(
    race_time_min ~ .,
    data = model_data[train_id, ],
    ntree = 500,
    importance = TRUE
  )

  pred <- predict(fit, newdata = model_data[-train_id, ])
  actual <- model_data$race_time_min[-train_id]
  c(
    RMSE = sqrt(mean((pred - actual)^2)),
    MAE = mean(abs(pred - actual)),
    R2 = cor(pred, actual)^2
  )
  randomForest::importance(fit)
} else {
  message("Install randomForest to run random forest regression.")
}
```

## k-nearest neighbors regression

```{r knn-regression}
data <- read_example("knnRegression.csv")
model_data <- subset(data, select = -participant)

if (has_pkg("caret")) {
  set.seed(1)
  ctrl <- caret::trainControl(method = "cv", number = 10)

  fit <- caret::train(
    wellbeing_score ~ sleep_hours + stress_score + physical_activity_min_week + social_support_score,
    data = model_data,
    method = "knn",
    trControl = ctrl,
    preProcess = c("center", "scale"),
    tuneLength = 10
  )

  fit
} else {
  message("Install caret to run k-nearest neighbors regression.")
}
```

# Machine learning classification

## Decision tree classifier

```{r decision-tree-classifier}
data <- read_example("decisionTreeClassifier.csv")
data$class <- factor(data$class)
model_data <- subset(data, select = -participant)

if (has_pkg("rpart")) {
  set.seed(1)
  train_id <- sample(seq_len(nrow(model_data)), size = floor(0.70 * nrow(model_data)))

  fit <- rpart::rpart(
    class ~ .,
    data = model_data[train_id, ],
    method = "class"
  )

  pred <- predict(fit, newdata = model_data[-train_id, ], type = "class")
  table(Predicted = pred, Observed = model_data$class[-train_id])
  mean(pred == model_data$class[-train_id])

  plot(fit)
  text(fit)
} else {
  message("Install rpart to run decision tree classification.")
}
```

## Random forest classifier

```{r random-forest-classifier}
data <- read_example("randomForestClassifier.csv")
data$class <- factor(data$class)
model_data <- subset(data, select = -athlete)

if (has_pkg("randomForest")) {
  set.seed(1)
  train_id <- sample(seq_len(nrow(model_data)), size = floor(0.70 * nrow(model_data)))

  fit <- randomForest::randomForest(
    class ~ .,
    data = model_data[train_id, ],
    ntree = 500,
    importance = TRUE
  )

  pred <- predict(fit, newdata = model_data[-train_id, ])
  table(Predicted = pred, Observed = model_data$class[-train_id])
  mean(pred == model_data$class[-train_id])
  randomForest::importance(fit)
} else {
  message("Install randomForest to run random forest classification.")
}
```

## k-nearest neighbors classifier

```{r knn-classifier}
data <- read_example("knnClassifier.csv")
data$class <- factor(data$class)
model_data <- subset(data, select = -participant)

if (has_pkg("caret")) {
  set.seed(1)
  ctrl <- caret::trainControl(method = "cv", number = 10)

  fit <- caret::train(
    class ~ .,
    data = model_data,
    method = "knn",
    trControl = ctrl,
    preProcess = c("center", "scale"),
    tuneLength = 10
  )

  fit
} else {
  message("Install caret to run k-nearest neighbors classification.")
}
```

## Naive Bayes classifier

```{r naive-bayes-classifier}
data <- read_example("naiveBayes.csv")
data$class <- factor(data$class)
model_data <- subset(data, select = -student)

if (has_pkg("e1071")) {
  set.seed(1)
  train_id <- sample(seq_len(nrow(model_data)), size = floor(0.70 * nrow(model_data)))

  fit <- e1071::naiveBayes(class ~ ., data = model_data[train_id, ])
  pred <- predict(fit, newdata = model_data[-train_id, ])

  table(Predicted = pred, Observed = model_data$class[-train_id])
  mean(pred == model_data$class[-train_id])
} else {
  message("Install e1071 to run naive Bayes classification.")
}
```

# Principal component analysis

## Principal component analysis

```{r principal-component-analysis}
data <- read_example("principalComponentAnalysis.csv")
variables <- c(
  "attention_accuracy",
  "working_memory",
  "processing_speed",
  "inhibition_score",
  "cognitive_flexibility",
  "verbal_memory",
  "spatial_reasoning"
)

vars <- data[, variables]
fit <- prcomp(vars, center = TRUE, scale. = TRUE)

summary(fit)
fit$rotation
biplot(fit, scale = 0)
```
